---
title: "LME_Workshop"
author: "R. Alex Thompson"
date: "2/27/2023"
output: pdf_document
---

First, we're going to import all of the libraries needed today, along with our toy data. Uncomment the lines if you need install the packages and remember to recomment them once installed. Otherwise, installation will occur every time you use this document.   
  
``` {r setup}

# install packages
#install.packages('dplyr')
#install.packages('ggplot2')
#install.packages('lme4')
#install.packages('gee')
#install.packages('ggeffects')

# load libraries
library(dplyr) # for manipulating our data
library(ggplot2) # for visualization
library(lme4) # for linear mixed models
library(gee)# for general estimation models
library(ggeffects) # for plotting the GEE results


# to read the data in, we want to set the working directory
# change this to your file path
setwd("/Users/alexthompson/Dropbox/PhD data/Stats stuff/Mixed Effects workshop 2023") 

setwd("C:/Users/robert.a.thompson/Dropbox/PhD data/Stats stuff/Mixed Effects workshop 2023") 



data<-read.csv("data.csv")

```
  

First, let's look at our data 

``` {r data overview}

head(data) # take a look at the data

# this plots all of the data on an x,y plane and fits a simple linear function to it
p1<-ggplot(data,aes(x,y)) + geom_point() + geom_smooth(method="lm") + theme_minimal()
p1

```

``` {r lm}
# let's estimate a linear model on the data and put numbers to what we see in the figure
summary(lm(y~x,data))

```


``` {r group level look}


p2<-ggplot(data,aes(x,y,color=group)) + geom_point() + geom_smooth(method="lm") + theme_minimal()
p2

summary(lm(y~x*group,data))
coef(aov(y~x*group,data))
summary(aov(y~x*group,data))
```


``` {r individual level look}
p3<-ggplot(data,aes(x,y,color=i)) + geom_point() + geom_smooth(method="lm") + theme_minimal()
p3

summary(lm(y~x*i,data))


```

  
Let's start with Linear Mixed Effects Models. INSERT A DISCUSSION OF THIS.  
  
``` {r LME intercept only}

# intercept only random model
mod1<-lmer(y~x+(1|i),data=data)
mod1
summary(mod1)

```

``` {r LME slope only}

# slope only random model
mod2<-lmer(y~x+(0 + x|i),data=data)
mod2
summary(mod2)

```

``` {r LME slope and intercept}
# slope and intercept random model
mod3<-lmer(y~x+(1 + x|i),data=data)
mod3
summary(mod3)
```


``` {r visualizing LME}
# visualizing LME
ggplot(data,aes(x,y)) + geom_point() +
  geom_smooth(method="lm",color="black",mapping=aes(y=predict(mod3,data))) + # this part plots the marginal linear model after accounting for all of the individual variation
  geom_smooth(method="lm",mapping=aes(x,y,color=i),linewidth=0.5) + # but we also want to show individual responses from our mixed effects model so we include a new mapping that incorporates individual-level as well
  theme_minimal()
```
  
A standard linear mixed effects model works well in the case of continuous data. What happens if we have count, or presence absence data? As with standard linear models, we can use a $generalized$ model.  
  
``` {r GLM read and view}
# read in data for GLM
countdata<-read.csv("count_data.csv")%>%
  mutate(Date=paste(1,Month,Year,sep="-"))
countdata$Date<-as.Date(countdata$Date,format="%d-%B-%Y")
head(countdata)

# take a quick look at the data
ggplot(countdata,aes(Date,Dead,color=Site,group=Site)) + geom_point() + geom_line()+ theme_minimal()


```

``` {r fit GLM1}
# let's fit a GLM to our data
# month as random effect
glm1<-glmer(Dead~Site+(1|Month),family="poisson",data=countdata)
summary(glm1)
```

``` {r fit GLM2}
# year as random effect
glm2<-glmer(Dead~Site+(1|Year),family="poisson",data=countdata)
summary(glm2)
```

``` {r fit GLM3}
# date (month and year) as random effect
glm3<-glmer(Dead~Site+(1|Date),family="poisson",data=countdata)
summary(glm3)
```

``` {r GLM viz pooled}
# visualizing GLMMs
ggplot(countdata,aes(Date,Dead,color=Site,group=Site)) + 
  geom_point() + 
  geom_line()+ 
  geom_smooth(method="glm")+
  theme_minimal()

```

``` {r GLM viz Year}

ggplot(countdata,aes(Date,Dead,color=Site,group=Year)) +
  geom_point() +
  geom_smooth(method="glm") +
  theme_minimal() +
  facet_wrap(~Site)

```
  
Now, a special case of mixed-effects models is a general estimation equation. Unlike a mixed-effects model where we get subject-level estimates, a GEE integrates out these subject-level variances and instead gives a population effect. 

``` {r GEE}
# make a general estimation equation
data<-arrange(data,i)
data$i<-as.factor(data$i)
gee_mod<-gee(y~x*group,
             data=data,
             id=i,
             family="gaussian",
             corstr="AR-M",Mv=1)
summary(gee_mod)
```
